Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-14 write-ups/PLDI/paper.tex
Edit figure caption: remove nb_queues
Polishing and shortening.

Working with cooperative kernels requires some manual effort on the
part of the programmer: they must place the $\transmit$ qualifier and
use the new language primitives described in
\mysec\ref{sec:semantics}, and account for slight changes in the OpenCL semantics induced by these new features.

The programming model and associated language constructs were inspired
by the requirements of the irregular GPU algorithms proposed in the
literature.  For these algorithms, it was usually easy to identify
where to place the constructs.

We discuss the main semantic change induced by our programming
model---that the number of workgroups executing a kernel varies
dynamically---and illustrate our programming model and the impact of
this semantic change by adapting the work stealing and graph traversal
algorithms of \mysec\ref{sec:openclexamples} to use cooperative
kernels.

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel: it may change due to workgroups executing
$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Because $\getglobalsize()$, which retrieves the total number of threads executing the kernel, is
a function of $\getnumgroups()$, the value it returns is also subject to change.
The programmer
must therefore write their cooperative kernel in a manner that is
robust to changes in the values returned by these functions.  In our experience this either involves avoiding the use of $\getnumgroups()$ at all, or writing a kernel where the only opportunity for the number of groups to change comes from $\resizingglobalbarrier()$, in which case the number of active groups does offer some stability, as discussed below.

@@

\myfiglong\ref{fig:workstealing-cooperative} shows a cooperative
version of the work stealing example of \myfig\ref{fig:workstealing} that we
discussed in \mysec\ref{sec:openclexamples}.

In this example there is no state to transmit since a computation is
entirely parameterised by a task, which is retrieved from a queue
located in global memory. In the main loop, before trying to obtain a
task, we make the workgroup offer itself to be forked or killed
(lines~\ref{line:cwksteal:offerfork},\ref{line:cwksteal:offerkill}). Note
that a workgroup may be killed even if its associated task queue is
not empty, since remaining tasks could eventually be stolen by other
workgroups.  Care must be taken on computing queue identifiers: the
original version in \myfig\ref{fig:workstealing} can rely on the fact
that there are as many queues as workgroups, but here the number of
workgroups may change during execution. Therefore, the number of
queues must now be explicitely given as an argument
(line~\ref{line:cwksteal:argnbqueue}) and used as an upper bound of
queue indexes when scanning queues in $\mathsf{more\_work}$
(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
workgroup's queue index (line~\ref{line:cwksteal:getgroupid}) must now
be moved inside the main loop since a forked workgroup that starts to
execute after $\offerfork$ must re-compute its group id.

@@

\paragraph{Cooperative kernels that only use barriers}

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  interval} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.

This guarantee can be exploited by algorithms that perform strided
data processing.  We now illustrate this using the graph traversal motivating example.

\paragraph{Adapting graph traversal} 
\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
graph traversal kernel of \myfig\ref{fig:graphsearch}, introduced in
\mysec\ref{sec:openclexamples}.  On lines~\ref{line:cgraph:resizing1}
and ~\ref{line:cgraph:resizing2}, we change the original global
barriers into a resizing barriers. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barriers
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so
that new workgroups know which frontier they are processing;
$\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so
that new workgroups know which of the node arrays are input and
output. Lastly, the static work distribution of the original kernel is
no longer valid in a cooperative kernel. This is because the stride
(which is based on $M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}). This example exploits the fact that
the cooperative kernel does not issue $\offerkill$ nor $\offerfork$
directly: the value of $\keyword{stride}$ obtained from
$\getglobalsize()$ at line~\ref{line:cgraph:rechunking2} is stable
until the next resizing barrier at line~\ref{line:cgraph:resizing1},
and in particular can be safely used inside the $\keyword{for}$ loop.

@@

In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
in our experiments, which largely captures the irregular blocking
algorithms that are available as open source GPU kernels.  These all
employ either work stealing or operate on graph data structures.
Like the simplified example of \myfig\ref{fig:workstealing-cooperative}, the work stealing algorithms have a transactional flavour
and require little or no state to be carried between transactions.  Thus the point at which a workgroup is ready to process a new task is a natural point to place $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.
%
The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
most level-by-level graph algorithms.  These algorithms tend to
require more state to be annotated with $\transmit$, and in some cases
a global barrier is required in the midst of a transaction, in which
case $\globalbarrier$, but not $\resizingglobalbarrier$, is suitable
for achieving synchronization without disturbing the set of active
workgroups.  But often it is the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.

In summary, our experience so far indicates that cooperative kernels
is a good fit for current irregular algorithms.  Of course this is, in
part, because we designed the model with these algorithms in mind, but
we argue that the fit is natural not just for the examples that we
study, but more generally for at least these two patterns of irregular
algorithm.

@@

The semantics presented in \mysec\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.
%
However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient execution} of long-running kernels
(e.g.\ implementing irregular data-parallel algorithms), and
\emph{responsive multi-tasking} so that the GPU can be shared between
one or more cooperative kernels and a number of short, high-priority
tasks with soft real-time constraints.

To achieve this balance, both an implementation of the cooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet the following non-functional requirements.

@@

The purpose of $\offerkill$ is to provide the scheduler with an
opportunity to destroy a workgroup in order to schedule
higher-priority tasks.  Because these higher-priority tasks, such as
graphics rendering, may have soft real-time constraints, the scheduler
relies on the cooperative kernel to execute $\offerkill$ sufficiently
frequently that the scheduler can meet such real-time constraints.

Using work stealing as an example: in the kernel of \myfig\ref{fig:workstealing-cooperative}, a
workgroup offers itself to the scheduler after processing each task.
If tasks turn out to be very short then it might turn out to be
reasonable to invoke $\offerkill$ less frequently, after every $k$
tasks for some appropriate $k$.  On the other hand, if individual
tasks may be very time-consuming then it may be necessary to rewrite
the algorithm so work on a larger number of shorter tasks, so as to
achieve a higher rate of calls to $\offerfork$.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide with
cooperative kernel with as many workgroups, up to this number, as
other constraints allow.  As a result, the scheduler should generally
accept an $\offerkill$ only if a compute unit is required for another
task, and should respond positively to $\offerfork$ calls by
scheduling new workgroups if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.

However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.  With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.

>>>>

We discuss the main semantic change induced by our programming
model---that the number of workgroups executing a kernel varies
dynamically---and illustrate the new language constructs and the
impact of this semantic change by adapting the work stealing and graph
traversal algorithms of \mysec\ref{sec:openclexamples} to use
cooperative kernels.

\paragraph{A changing number of workgroups}  Unlike in regular OpenCL,
the value returned by $\getnumgroups()$ is not fixed during the
lifetime of a cooperative kernel: it corresponds to the active group count $M$ and may change due to workgroups executing
$\offerkill$, $\offerfork$ and $\resizingglobalbarrier$.  Being a function of $\getnumgroups()$, the value returned by $\getglobalsize()$
is also subject to change.
A cooperative kernel must therefore be written in a manner that is
robust to changes in the values returned by these functions.

In general, their volatility means that use of these functions should be avoided.
However, the situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  interval} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.
%
This guarantee can be exploited by algorithms that perform strided
data processing, which we illustrate below using the graph traversal
example.

@@

\myfiglong\ref{fig:workstealing-cooperative} shows a cooperative
version of the work stealing example of \myfig\ref{fig:workstealing} from \mysec\ref{sec:openclexamples}.
%
In this example there is no state to transmit since a computation is
entirely parameterised by a task, which is retrieved from a queue
located in global memory. In the main loop, before trying to obtain a
task, we make the workgroup offer itself to be forked or killed
(lines~\ref{line:cwksteal:offerfork}-\ref{line:cwksteal:offerkill}). Note
that a workgroup may be killed even if its associated task queue is
not empty, since the remaining tasks will be stolen by other
workgroups.  Care must be taken on computing queue identifiers: the
original version in \myfig\ref{fig:workstealing} can rely on the fact
that there are as many queues as workgroups, but here the number of
workgroups may change during execution. Therefore, the number of
queues must now be explicitely given as an argument
(line~\ref{line:cwksteal:argnbqueue}) and used as an upper bound of
queue indexes when scanning queues in $\mathsf{more\_work}$
(line~\ref{line:cwksteal:morework}) and $\mathsf{pop\_or\_steal}$
(line~\ref{line:cwksteal:poporsteal}). Besides, computation of the
workgroup's queue index (line~\ref{line:cwksteal:getgroupid}) must now
be moved inside the main loop since a forked workgroup that starts to
execute after $\offerfork$ must re-compute its group id.

@@

\paragraph{Adapting graph traversal} 
\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the
graph traversal kernel of \myfig\ref{fig:graphsearch} from
\mysec\ref{sec:openclexamples}.  On lines~\ref{line:cgraph:resizing1}
and ~\ref{line:cgraph:resizing2}, we change the original global
barriers into a resizing barriers. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barriers
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so
that new workgroups know which frontier they are processing;
$\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so
that new workgroups know which of the node arrays to use for input and
output. Lastly, the static work distribution of the original kernel is
no longer valid in a cooperative kernel. This is because the stride
(which is based on $M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}). This example exploits the fact that
the cooperative kernel does not issue $\offerkill$ nor $\offerfork$
directly: the value of $\keyword{stride}$ obtained from
$\getglobalsize()$ at line~\ref{line:cgraph:rechunking2} is stable
until the next resizing barrier at line~\ref{line:cgraph:resizing1},
and in particular can be safely used inside the $\keyword{for}$ loop.

@@

\paragraph{Patterns for irregular algorithms}
In \mysec\ref{sec:portingalgorithms} we describe the set of irregular GPU algorithms used
in our experiments, which largely captures the irregular blocking
algorithms that are available as open source GPU kernels.  These all
employ either work stealing or operate on graph data structures, and placing our new constructs follows a common, easy-to-follow pattern in each case.

Like the simplified example of \myfig\ref{fig:workstealing-cooperative}, the work stealing algorithms have a transactional flavour
and require little or no state to be carried between transactions.  Thus the point at which a workgroup is ready to process a new task is a natural point to place $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.

The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
most level-by-level graph algorithms.  These algorithms tend to
require some state to be transmitted, and in some cases
a global barrier is required in the midst of a transaction, in which
case $\globalbarrier$, but not $\resizingglobalbarrier$, is suitable
for achieving synchronization without disturbing the set of active
workgroups.  But often it is the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.

Our practical experience suggests that cooperative kernels are a good
fit for at least these two classes of irregular algorithms, which is
of course partly because we were inspired by these algorithms in
designing the model.

@@

The semantics presented in \mysec\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.
%
However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient} execution of algorithms that require fair scheduling, and
\emph{responsive} multi-tasking, so that the GPU can be shared between
cooperative kernels and other shorter tasks with soft real-time constraints.
%
To achieve this balance, an implementation of the cooperative
kernels model, and the programmer of a cooperative kernel, must strive
to meet the following non-functional requirements.

@@

The purpose of $\offerkill$ is to provide the scheduler with an
opportunity to destroy a workgroup in order to schedule
higher-priority tasks.  The scheduler relies on the cooperative kernel
to execute $\offerkill$ sufficiently frequently that soft real-time
constraints of other workloads can be met.
%
Using our work stealing example: in
\myfig\ref{fig:workstealing-cooperative} a workgroup offers itself to
the scheduler after processing each task.  If tasks are sufficiently
fast to process then the scheduler will have ample opportunities to
de-schedule workgroups.  But if tasks can be very time-consuming to
process then it might be necessary to rewrite the algorithm so that
tasks are shorter and more numerous, to achieve a higher rate of calls
to $\offerkill$.

@@

Recall that the cooperative kernel would ideally utilise the $N$
workgroups requested upon kernel launch, in order to conduct its
processing efficiently.  The scheduler should thus aim to provide the
cooperative kernel $N$ workgroups if other constraints allow it,
by accepting an $\offerkill$ only if a compute unit is required for another
task, and responding positively to $\offerfork$ calls if compute units are available.  Our
implementation of cooperative scheduling (\mysec\ref{sec:implementation}) is as
generous as possible to the cooperative kernel in this regard.

\ADComment{This could be shorter.}
However, these are not hard-and-fast rules: a scheduler might employ
heuristics based on historical workloads to avoid thrashing.  For
example, if graphics calls are known to arrive at a certain rate then
the scheduler might decide that it is worth accepting an $\offerkill$
from the cooperative kernel on the assumption that a graphics task
will be requested momentarily.  Similarly, the scheduler might provide
some, but not all, compute units to the cooperative kernel in response
to an $\offerfork$ call if it has reason to believe that the remaining
compute units are likely to be useful for other high priority tasks
that are likely to arrive imminently.  With some additional language extensions, it would be possible for the cooperative kernel to hint to the scheduler if, for example, the kernel is at a point of computation where there is less work to do than usual, so that $\offerkill$ calls could be accepted in order to save energy.
